Expert-driven approximation of MPE

### 2019-08-06 literature-review/explaining-mpe.tex
Literature Review

\section{Explaining the Most Probable Explanation}\label{sec:explaining-mpe}
\cite{Butz2018}

>>>>

\section{Explaining the Most Probable Explanation} \label{sec:explaining-the-most-probable-explanation}
The paper \enquote{Explaining the Most Probable Explanation} by \cite{Butz2018} inserts itself into the literature concerned with the explainability of Bayesian Networks.
In particular, taking \cite{lacave2002review}'s classification presented in Sec. \ref{sec:explainability-in-bayesian-networks}, it attempts to define a \textit{verbal explanation} to the \textit{knowledge base} and of the \textit{propagated evidence}.
Unlike \cite{lacave2002review}'s definition of explanation of a knowledge base and of other previous works, the paper is not concerned with finding the most probable assignment of variables that would explain the given evidence but, rather, the inverse problem.
By starting with evidence and finding a maximally probable configuration, the authors hope \enquote{to look at the complete scenario to get an overview before deciding which variables should be focused on}; that is, the goal appears to give the user an overview of the situation.

The initial claim of the paper is that even though BNs provide a graphical structure to the knowledge base, they remain of difficult interpretation for domain experts.
The examples brought to justify the claim are that edges in the graph do not necessarily represent causal dependencies and that d-separation (defined in \ref{subsec:d-separation}) may be confusing.
The authors plan to address this claim by constructing a \textit{dialogue} with the user and thus to continue in the long tradition of dialogical approaches to explaining BNs.

The defining characteristic of their approach is that the domain expert is able to \enquote{argue} with the MPE and investigate alternative explanations.
The complete methodology, that is executed over three steps, is shown in Fig. \ref{fig:butz-methodology}.
The first step is the construction of the \enquote{knowledge base}, which is nothing else than a probability tree representing a \enquote{chain of deduction} constructed following the strongest probabilistic dependencies between variables in the BN.
Such a knowledge base is convenient because the document plan for the Natural Language Generation step is directly derived from it.
One issue that is immediately apparent is that this greedy approach does not \enquote{generate the MPE solution} as the authors claim, as noted in Subsec. \ref{subsec:bnupdating}.
This does not discredit the argumentative method as a whole, as it is not necessary for the user to be arguing the MPE to derive a good explanation; this ties into one of the main findings in the previous sections that many xAI researchers are only focusing on half of what an explanation is.
A good explanation is not given only by its formal properties but, most importantly, by how it acts as an interface between the real \textit{user} and the model.
This is what \cite{abdul2018trends} mean when they lament that \enquote{despite their mathematical rigor, these works \textit{[referring to the existing explainability methods]} suffer from a lack of usability, practical interpretability and efficacy on real users.}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{literature-review/images/butz-methodology}}
\caption{\cite{Butz2018}}
\label{fig:butz-methodology}
\end{figure}

The document plan for the argumentation follows the same chain of strongest dependencies constructed in the knowledge base until the expert disagrees; at that point, the user is presented with an alternative \enquote{MPE}.
An example of how the document plan may look after interaction with the user is shown in Fig. \ref{fig:butz-tree}.
All the natural language phrasing is generated via boilerplates that realise both the micro-planning phase and the generation of the text.

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{literature-review/images/butz-tree}}
\caption{\cite{Butz2018}}
\label{fig:butz-tree}
\end{figure}

The authors recognise that such chains of deduction could become long and cognitively overloading in the case of larger BNs, as every variable in the tree is explained by all its ancestors.
A solution they propose is that of \textit{pruning} the probability tree by excluding d-separated nodes and those under a certain threshold of significance.
They also adapt some methods from literature to perform \textit{conflict analysis}; that is, only variables that contribute positively to the explanation are maintained in the document plan.

On the whole, the paper offers a compelling explanation method for BNs by building on an established tradition of explainability through dialogue.
The work, though, takes some methodological missteps and also continues the \enquote{sin} of not validating its claims on real users, that is one of the primary gaps in the xAI field that were identified in the previous sections.

### 2019-08-06 mathematical-background/probability-theory.tex
Literature Review

\subsection{Probability distributions} \label{subsec:probability-distributions}
\begin{definition}
	A \textit{probability distribution} is a function $\mathbb{P}: \mathcal{S} \rightarrow \mathbb{R}$ with $\mathcal{S}$ a set of \textit{events} of interest.  
To be a valid probability distribution $\mathbb{P}$ must satisfy:
\begin{itemize}
	\item $\mathbb{P}(\sigma) \geq 0 \quad \forall \sigma \in \mathcal{S}$
	\item $\sum_{\sigma} = 1 \quad \forall \sigma \in \mathcal{S}$
	\item $\alpha, \beta \in \mathcal{S} \wedge \alpha \cap \beta=\emptyset 	\Rightarrow \mathbb{P}(\alpha \cup \beta)=\mathbb{P}(\alpha)+\mathbb{P}(\beta)$
\end{itemize}
\end{definition}
Each event $\sigma \in \mathcal{S}$ must have a probability $\mathbb{P}(\sigma) \in [0,1]$ and the sum of all these must equal $1$. 
An event with $P(\sigma) = 0$ is deemed \textit{impossible} while one with $\mathbb{P}(\sigma) = 1$ is \textit{certain}.

@@

Equation \ref{eq:conditionalprobability} can be easily manipulated to obtain another basic element of Bayesian Networks: what is called the \textit{chain rule of conditional probabilities}:
\begin{equation} \label{eq:chainrule}
	\mathbb{P}(\beta \cap \alpha) = \mathbb{P}(\beta \mid \alpha) \mathbb{P}(\alpha)
\end{equation}
This can be generalised to any number of events:
\begin{equation} \label{eq:chainrule}
	\mathbb{P}(\alpha_1 \cap \ldots \cap \alpha_n) = \mathbb{P}(\alpha_n \mid \alpha_1 \cap \ldots \cap \alpha_{n-1}) \ldots \mathbb{P}(\alpha_1 \mid \alpha_2 ) \mathbb{P}(\alpha_1) 
\end{equation}
Intuitively, it means that we can decompose joint probabilities as products of conditional probabilities.  
As we'll see, this is how the values in a Bayesian Network are calculated.

>>>>

\subsection{Probability Distributions} \label{subsec:probability-distributions}
\begin{definition}
	A \textit{probability distribution} is a function $\mathbb{P}: \mathcal{S} \rightarrow \mathbb{R}$ with $\mathcal{S}$ a set of \textit{events} of interest.  
To be a valid probability distribution $\mathbb{P}$ must satisfy:
\begin{itemize}
	\item $\mathbb{P}(\sigma) \geq 0 \quad \forall \sigma \in \mathcal{S}$
	\item $\sum_{\sigma} = 1 \quad \forall \sigma \in \mathcal{S}$
	\item $\alpha, \beta \in \mathcal{S} \wedge \alpha \cap \beta=\emptyset 	\Rightarrow \mathbb{P}(\alpha \cup \beta)=\mathbb{P}(\alpha)+\mathbb{P}(\beta)$
\end{itemize}
\end{definition}
Each event $\sigma \in \mathcal{S}$ must have a probability $\mathbb{P}(\sigma) \in [0,1]$ and the sum of all these must equal $1$. 
An event with $P(\sigma) = 0$ is deemed \textit{impossible} while one with $\mathbb{P}(\sigma) = 1$ is \textit{certain}.

@@

Equation \ref{eq:conditionalprobability} can be easily manipulated to obtain another basic element of Bayesian Networks: what is called the \textit{chain rule of conditional probabilities}:
\begin{equation} \label{eq:chainrule}
	\mathbb{P}(\beta \cap \alpha) = \mathbb{P}(\beta \mid \alpha) \mathbb{P}(\alpha)
\end{equation}
This can be generalised to any number of events:
\begin{equation} \label{eq:chainrule}
	\mathbb{P}(\alpha_1 \cap \ldots \cap \alpha_n) = \mathbb{P}(\alpha_n \mid \alpha_1 \cap \ldots \cap \alpha_{n-1}) \ldots \mathbb{P}(\alpha_1 \mid \alpha_2 ) \mathbb{P}(\alpha_1) 
\end{equation}
Intuitively, it means that we can decompose joint probabilities as products of conditional probabilities.  
This is how the probability values in a Bayesian Network are calculated.

Another immediate, and crucial, consequence of Eq. \ref{eq:conditionalprobability} is known as \textit{Bayes' Theorem}, that lets us calculate the probability of event $\alpha$ given our knowledge of event $\beta$.
\begin{definition} \label{eq:bayes-theorem}
	Bayes' Theorem states that the probability of even $\alpha$ conditional on even $\beta$ is given by:
	\begin{equation}
		\mathbb{P}(\alpha | \beta)=\frac{\mathbb{P}(\beta | \alpha) \mathbb{P}(\alpha)}{\mathbb{P}(\beta)}
	\end{equation}
\end{definition}
Intuitively, this describes the probability of an event, based on some prior knowledge of another event.
For example, our belief in a person having breast cancer ($\alpha$) is increased if we know that the person is older than 50 years of age ($\beta$).

### 2019-08-06 methodology/data-set.tex
mutual information section
Literature Review

% !TEX root = thesis-thomas-tiotto.tex

\section{Data set} \label{sec:data-set}
As anticipated in Chap. \ref{chap:introduction} the work carried out in this thesis had a certain degree of collaboration with a third party, \cite{istitutocantonalepresentazione}.

\subsection{Istituto Cantonale di Patologia}
Istituto Cantonale di Patologia (ICP) is an institute based in Locarno that is specialised in the histological analysis of tissue samples received from private patients, clinics and hospitals, mainly in support of cancer diagnosis.
Its Laboratory of Medical Diagnostics supports pathologists in the diagnosis of neoplastic diseases through the application of cytogenetic techniques; that is, the focus is on understanding how chromosomes relate to cell behaviour, particularly during mitosis and meiosis.
One of the techniques used is Fluorescence in Situ Hybridization (FISH) that is able to localise the presence or absence of specific DNA sequences in chromosomes.
These tests are aimed at identifying the precise profile of the cancer cells and thus inform the clinician on the best treatment for the specific patient.

In addition to its clinical support activities, the Istituto also carries out scientific research aimed at better understanding certain types of cancers at a basic level.
In the last ten years, the ICP has published more than 200 peer-reviewed papers and more than 100 works in non-peer reviewed journals and is active at a national and international level.

\subsection{Motivation}
My first contact with the ICP was during a meeting with Dr. Vittoria Martin (\cite{martin2012}), molecular citogenetist, in date 28/01/2019.
The institute had expressed interest in bringing machine learning into their workflow in order to both augment their profiling capabilities for patients and to be able to extract new knowledge from their existing data.
This knowledge-extraction may lead towards the confirmation of current scientific theories or may be the first step towards the formulation of novel ones.

\todo[inline]{chiedere a Vittoria cosa si aspettavano/aspettano}

My interest in collaborating with the Istituto stemmed from the desire to apply the methods described in Sec. \ref{sec:novel-contributions} to a real-world case.
Being the theoretical work being carried out in this thesis an expert-driven MPE approximation, collaboration with the institute has also provided the opportunity to implement a proof of concept using real histological data.
The doctors and researchers of the Istituto have been able to validate, from an Explainable AI and clinical relevance points of view, the model software that I have developed.
That is to say, they have validated the capacity of the developed software both in its capacity to support clinical decision making and to surface clarifying explanations of the data set and in its adherence to established medical literature.

@@

\subsection{Provided Data Set}
The data set I was provided with was created by \textit{Registro Tumori Ticino} (Locarno, Ticino) in order to highlight possible new relations between clinical, histopathological and molecular features, as well as to potentially discover novel biomarkers for the progression of the disease 
It consists of the histological records over 38 recorded variables of 3218 breast cancer patients who have been diagnosed between the years 2005 and 2014 within the Ticino canton of Switzerland.
The data set had been pre-processed by collaborators of IDSIA in agreement with the ICT with 13 of the variables being dropped, because not considered relevant.
In particular, all variables relating to the patient post-treatment were discarded as well as those recording the diagnosis date.
In Tab. \ref{tab:datasetvariables} is a description of the measured variables, together with their clinical meaning.
The value distribution of the data set is shown in Tab. \ref{tab:datasetdistribution}.

@@

\begin{table*}[htbp]
\caption{Data set preprocessing steps}
\begin{tabularx}{\textwidth}{@{} l Y @{}}
\toprule 
\textbf{Variable} & Action \\
\midrule 
\textbf{Codice globale} & Remove variable \\
\textbf{mut17q21} & Remove blanks \\
\textbf{loss 17} & Remove blanks \\
\textbf{eta arrotondata} & Bin into \enquote{$< 40$}, \enquote{$40-50$}, \enquote{$\geq 50$} \\
\textbf{lateralita} & Remove blanks and \enquote{sconosciuta} \\
\textbf{situ} & Remove blanks \\ \addlinespace
\textbf{morfologia} & Remove blanks and \enquote{unuseful} if performance on classification is subpar \\ \addlinespace
\textbf{pT} & Remove blanks and \enquote{unuseful}  \\
\textbf{pN} & Remove blanks and bin into \enquote{0} and \enquote{$\neq0$}\\
\textbf{M} & Remove blanks \\ 
\textbf{differenziazione} & Remove blanks and \enquote{Sconosciuto o non applicabile} \\ \addlinespace
\textbf{recettori estrogeni} & Remove blanks and bin into \enquote{negativo} if $\leq 10$,
		\enquote{debolmente positivo} if $\leq 50$, 
		\enquote{fortemente positivo} if $> 50$ \\ \addlinespace
\textbf{recettori progestinici} & Remove blanks and bin into \enquote{negativo} if $\leq 10$, 
		\enquote{debolmente positivo} if $\leq 50$, 
		\enquote{fortemente positivo} if $> 50$ \\ \addlinespace
\textbf{c erbB 2} & Remove blanks \\ 
\textbf{ki67} & Remove blanks and bin into \enquote{<14}, 
		\enquote{14-20}, \enquote{20-30}, \enquote{>30} \\ 
\textbf{FISH} & Remove blanks \\
\bottomrule
\end{tabularx}
\label{tab:datasetpreprocess}
\end{table*}

>>>>

\section{Data set} \label{sec:data-set}

\subsection{Istituto Cantonale di Patologia} \label{subsec:istituto-cantonale}
Istituto Cantonale di Patologia (ICP) is an institute based in Locarno that is specialised in the histological analysis of tissue samples received from private patients, clinics and hospitals, mainly in support of cancer diagnosis.
Its Laboratory of Molecular Pathology supports the diagnostic approach to neoplastic diseases through the application of biomolecular and cytogenetics techniques, which focus on understanding the impact of molecular alterations in carcinogenesis.
One of the methods used is Fluorescence in Situ Hybridization (FISH) that is able to localise the presence or absence of specific DNA sequences in chromosomes.
These tests are aimed at identifying the precise profile of the cancer cells and thus inform the clinician on the best treatment for the specific patient.

Molecular diagnosis is the most recent approach in pathology. It mainly aims to support pathologists in the diagnosis of neoplastic diseases through the employment of molecular biology and cytogenetics techniques and it enables the deep understanding and monitoring of patients' profiles, as well as as defining the prognosis and ad-hoc therapies. All together, it allows the development of an increasingly personalised medicine, also in terms of research (e.g. understanding of pathological mechanisms and thus the possibilities of intervention).
However, it is still limited in its capabilities by the presence of a dishomogeneous information pool, especially in terms of knowledge extraction; a non-uninform data set entails, for the pathologist, an increased cognitive complexity because of having to manage a high number of variables and heterogeneous meaning behind the data; these consequently force to resort to problem decomposition and reduction.

In addition to its clinical support activities, the ICP also carries out scientific research aimed at better understanding certain types of cancers at translational level i.e., at the level of protein synthesis.
In the last ten years, the ICP has published more than 200 peer-reviewed papers and more than 100 works in non-peer reviewed journals and is active at a national and international level.

\subsection{Motivation} \label{subsec:motivation}
\todo{magari togliere CV e mettere descrizione}
The Istituto Cantonale di Patologia already had a collaboration in place with the Dalle Molle Institute for Artificial Intelligence (IDSIA) (\cite{idsia}) to investigate a series of specific issues, whose details are outside of the scope of this thesis.
The work carried out in this thesis was deemed of interest because its extended beyond the existing collaboration.
The Institute had originally expressed interest in bringing machine learning into their workflow in order to both augment its profiling capabilities for patients and to be able to extract new knowledge from their existing data; this was paired with an attention for more experimental research directions, as the facilitation of human-machine interaction.

Given that the theoretical work carried out in this thesis is an \textit{expert-driven MPE approximation}, collaboration with the Institute provided the precious opportunity to implement a proof-of-concept system (described in Chap. \ref{chap:methodology} and \ref{chap:results}) based on a real medical data set.
The first contact with the ICP was during a meeting with Dr. Vittoria Martin (\cite{martin2012}), molecular citogenetist, in date 28/01/2019.
Since then, the clinicians and researchers of the ICP have been able to validate, from an Explainable AI and clinical relevance point of view, the model software that has been developed.
That is to say, they have validated, to an extent that will be made clear in Chap. \ref{chap:results}, the capabilities of the developed software both in its capacity to support clinical decision making and to surface clarifying explanations of the data set and in the adherence of its outputs to established medical literature.
This is of great interest because the lack of real-world validation of ML systems with real domain experts is one of the prime gaps in the existing xAI literature (as discussed in depth in Chap. \ref{chap:literature-review}).

An example application for a clinician of the ICT, would be ability to \enquote{fill in the blanks} of a patient's profile, as it is not uncommon, for a variety of reasons, to have missing data.
This may be because of degraded or insufficient tissue samples or because some test may not yet be part of the standard diagnostic procedure, even though their importance may already be suggested by clinical research.
In other cases, patients may be missing a result because the specific test hadn't yet been invented, for example FISH wasn’t available prior to 2010, so an a-posteriori inference could be made possible.
Another crucial use may be in understanding and quantifying in an efficacious manner the relationship between clinical variables.
It is not uncommon for some variables to have been observed but for their clinical relevance not to have yet been determined; learning their relationship with other variables could potentially not only help in defining their importance in tracing new patient profiles in terms of diagnostic, prognosis and support in decision making but also in placing them in terms of pathological mechanisms.
These are all examples highlighting the importance of the \textit{inference} capabilities of a Machine Learning system, but my current work aims to address the interfacing of the human user with the software. \todo{aspettare idee Ginevra su ulteriori applicazioni cliniche}
It is also hoped that facilitating the process of knowledge-extraction may lead towards the confirmation of current scientific theories or may be the first step towards the formulation of novel ones.

@@

\subsection{Provided Data Set}
The data set I was provided with was created by \textit{Registro Tumori Ticino} (Locarno, Ticino) in order to highlight possible new relations between clinical, histopathological and molecular features, as well as to potentially discover novel biomarkers involved in the progression of the disease.
It consists of the histopathological records over 38 variables of interest of 3218 breast cancer patients who have been diagnosed between the years 2005 and 2014 within the Ticino canton of Switzerland.
The data set had been pre-processed by collaborators of IDSIA in agreement with the ICP with 13 of the variables being dropped, because not considered relevant.
In particular, all variables relating to the patient post-treatment were discarded as well as those recording the diagnosis date.
In Tab. \ref{tab:datasetvariables} is a description of the measured variables, together with their clinical meaning.
The value distribution of the data set is shown in Tab. \ref{tab:datasetdistribution}.

@@

\begin{table*}[htbp]
\caption{Data set preprocessing steps}
\begin{tabularx}{\textwidth}{@{} l Y @{}}
\toprule 
\textbf{Variable} & Action \\
\midrule 
\textbf{Codice globale} & Remove variable \\
\textbf{mut17q21} & Remove variable \\
\textbf{loss 17} & Remove variable \\
\textbf{eta arrotondata} & Bin into \enquote{$< 40$}, \enquote{$40-50$}, \enquote{$\geq 50$} \\
\textbf{lateralita} & Remove blanks and \enquote{sconosciuta} \\
\textbf{situ} & Remove blanks \\ \addlinespace
\textbf{morfologia} & Remove blanks and \enquote{unuseful} if performance on classification is subpar \\ \addlinespace
\textbf{pT} & Remove blanks and \enquote{unuseful}  \\
\textbf{pN} & Remove blanks and bin into \enquote{0} and \enquote{$\neq0$}\\
\textbf{M} & Remove blanks \\ 
\textbf{differenziazione} & Remove blanks and \enquote{Sconosciuto o non applicabile} \\ \addlinespace
\textbf{recettori estrogeni} & Remove blanks and bin into \enquote{negativo} if $\leq 10$,
		\enquote{debolmente positivo} if $\leq 50$, 
		\enquote{fortemente positivo} if $> 50$ \\ \addlinespace
\textbf{recettori progestinici} & Remove blanks and bin into \enquote{negativo} if $\leq 10$, 
		\enquote{debolmente positivo} if $\leq 50$, 
		\enquote{fortemente positivo} if $> 50$ \\ \addlinespace
\textbf{c erbB 2} & Remove blanks \\ 
\textbf{ki67} & Remove blanks and bin into \enquote{<14}, 
		\enquote{14-20}, \enquote{20-30}, \enquote{>30} \\ 
\textbf{FISH} & Remove variable \\
\bottomrule
\end{tabularx}
\label{tab:datasetpreprocess}
\end{table*}

### 2019-08-06 methodology/instruments.tex
Literature Review

\subsection{Libraries}\label{subsec:libraries}
\subsubsection{Pomegranate}
\textit{pomegranate} (\cite{pomegranate}) is an open-source probabilistic models package for python.
Its core philosophy is that every probabilistic model, from Hidden Markov to Bayesian Network, can be seen as a probability distribution and, as such, can be flexibly composed into hierarchical mixture models (\cite{Schreiber2017}).
The package implements:
\begin{itemize}
	\item Probability Distributions
	\item General Mixture Models
	\item Hidden Markov Models
	\item Bayes Classifiers and Na{\"i}ve Bayes
	\item Markov Chains
	\item \textbf{Bayesian Networks}
	\item Factor Graphs
\end{itemize}

@@

\subsection{Algorithms} \todo{thresholded dialogue aggiungere}
\todo{quanti dettagli implementativi devo mettere?  perchè ho un sacco di funzioni di visualizzazione etc}

\subsubsection{Model construction}
The data was given in \texttt{.xlsx} format and was imported using \texttt{panda}'s \texttt{read\_excel} function that returned a \texttt{DataFrame} object.
The so imported data was then preprocessed by dropping unwanted records and binning the remaining ones as described in Tab. \ref{tab:datasetpreprocess} .
The actual BN representation is learned at runtime by calling the \texttt{from\_samples} method of \texttt{pomegranate}'s \texttt{BayesianNetwork} to solve the structure learning problem (defined in Subsec. \ref{subsec:bnstructurelearning}).

\subsubsection{d-separation}
A na{\"i}ve implementation according to the definition (presented in Subsec. \ref{subsec:d-separation}) to check for d-separation between node $X$ and $Y$ would have a complexity in the order of the number of trails between $X$ and $Y$; this leads to an exponential in the size of the graph running time.
Luckily, \cite{koller2007dseparation} present what is a linear time algorithm to solve the problem.

The \texttt{reachable} procedure takes as input the DAG representing the Bayesian Network $\mathcal{G}$, a source variable $X$ and a set of observed variables $Z$; on exit it returns the set of variables $R$ that are reachable from $X$.
The procedure runs in two phases, traversing the graph twice: first bottom-up from leaves to roots, then viceversa.
During the first run, the algorithm finds all nodes $A$ that are ancestors of the evidence set $Z$.
During the second, the procedure distinguishes the direction it visits each node in order to determine if it is traversable or not.
Any node $Y$ that is not in the evidence set is marked as reachable; if it is being visited in direction \enquote{up} it can be traversed as the v-structure is a chain.
All the parents of $Y$ are marked to be visited in the \enquote{up} direction (i.e. from below) and the converse is done for $Y$'s children.
If $Y$ is being visited in the \enquote{down} direction its children are again added to be visited in the \enquote{down} direction, because $Y$ is traversable.
Additionally, if $Y$ happened to be in the set $A$, found in the first step, then $Y$'s parents are marked to be visited in the \enquote{up} direction then the collider is active and $Y$ can be traversed (a collider is open iff. the central node or any of its descendants are observed).

The full procedure can be found in \cite{koller2007dseparation}; my implementation shown in Alg. \ref{alg:d-separation} follows this pseudocode very closely but the procedure \texttt{d-separated}, instead of finding all nodes $R$ that are d-connected to the input $X$, tests if a given target $Y$ is d-separated from $X$ or not.
This gives some extra flexibility in how the function can be used.
To find the set $S$ of all nodes d-separated from $X$ I simply iterate the \texttt{d-separated} test over all nodes $V$ in the graph representing the BN.

@@

\subsubsection{MPE}
\todo{descrivere come ottenere MPE con Gurobi: exporter, UAI ecc...}

\subsubsection{Other Machine Learning methods}

>>>>

\subsection{Libraries}\label{subsec:libraries}
\subsubsection{Pomegranate}
\textit{Pomegranate} (\cite{pomegranate}) is an open-source probabilistic models package for Python.
Its core philosophy is that every probabilistic model, from Hidden Markov to Bayesian Network, can be seen as a probability distribution and, as such, can be flexibly composed into hierarchical mixture models (\cite{Schreiber2017}).
The package implements:
\begin{itemize}
	\item Probability Distributions
	\item General Mixture Models
	\item Hidden Markov Models
	\item Bayes Classifiers and Na{\"i}ve Bayes
	\item Markov Chains
	\item \textbf{Bayesian Networks}
	\item Factor Graphs
\end{itemize}

@@

\subsection{Algorithms} \label{subsec:algorithms}
\todo{quanti dettagli implementativi devo mettere?  perchè ho un sacco di funzioni di visualizzazione etc}

\subsubsection{Model construction}
The data was given in \texttt{.xlsx} format and was imported using \texttt{panda}'s \texttt{read\_excel} function that returned a \texttt{DataFrame} object.
The imported data was then preprocessed by dropping unwanted records and binning the remaining ones as described in Tab. \ref{tab:datasetpreprocess} .
The actual BN representation is learned at runtime by calling the \texttt{from\_samples} method of \texttt{pomegranate}'s \texttt{BayesianNetwork} to solve the structure learning problem (defined in Subsec. \ref{subsec:bnstructurelearning}).

The binned data was codified into integer representations before being passed to pomegranate's structure learning algorithm.
Thus the network's state names are in natural language but the internal representation of the values of each random variable is an integer number.
A dictionary object is used to translate one representation into the other at runtime.

\subsubsection{d-separation}
A na{\"i}ve implementation according to the definition (presented in Subsec. \ref{subsec:d-separation}) to check for d-separation between node $X$ and $Y$ would have a complexity in the order of the number of trails between $X$ and $Y$; this would lead to an exponential in the size of the graph running time.
Luckily, \cite{koller2007dseparation} present a linear time algorithm to solve the problem.

The \texttt{reachable} procedure takes as input the DAG representing the Bayesian Network $\mathcal{G}$, a source variable $X$ and a set of observed variables $Z$; on exit it returns the set of variables $R$ that are reachable from $X$.
The procedure runs in two phases, traversing the graph twice: first bottom-up from leaves to roots, then viceversa.
During the first run, the algorithm finds all nodes $A$ that are ancestors of the evidence set $Z$.
During the second, the procedure distinguishes the direction it visits each node in order to determine if it is traversable or not.
Any node $Y$ that is not in the evidence set is marked as reachable; if it is being visited in direction \enquote{up} it can be traversed as the v-structure is a chain.
All the parents of $Y$ are marked to be visited in the \enquote{up} direction (i.e. from below) and the converse is done for $Y$'s children.
If $Y$ is being visited in the \enquote{down} direction its children are again added to be visited in the \enquote{down} direction, because $Y$ is traversable.
Additionally, if $Y$ happened to be in the set $A$, found in the first step, then $Y$'s parents are marked to be visited in the \enquote{up} direction because the collider is active and $Y$ can be traversed (a collider is open iff. the central node or any of its descendants are observed).

The full procedure can be found in \cite{koller2007dseparation}; my implementation follows this pseudocode very closely but the procedure \texttt{d-separated}, instead of finding all nodes $R$ that are d-connected to the input $X$, tests if a given target $Y$ is d-separated from $X$ or not, as is shown in Alg. \ref{alg:d-separation}.
This gives some extra flexibility in how the function can be used.
To find the set $S$ of all nodes d-separated from $X$ I simply iterate the \texttt{d-separated} test over all nodes $V$ in the graph representing the BN.

@@

\subsubsection{MPE}
The solution to the Most Probable Explanation problem (MPE), as defined in Subsec. \ref{subsec:bnupdating}, is found by using DAOOPT (Subsec. \ref{subsec:libraries} ) as an external solver.
The latest version of DAOOPT was downloaded from the github page (\cite{daoopt}) and compiled into a executable.
DAOOPT only offers a command line interface so some extra work was needed in order to interface it with the Python-based application I have developed.
The connection was done by first writing to stable storage a \texttt{.uai} containing the model definition and a \texttt{.uai.evid} with the observed evidence.
These files are then piped to DAOOPT by using Python's \texttt{subprocess} module to run the following command in an external background shell:
\begin{verbatim}
	./daoopt -f pomegranate.uai -e pomegranate.uai.evid
\end{verbatim}
The shell output is captured and also written to stable storage, in order to be decoded and used in the system.

To exemplify the process, we return to the example used while presenting DAOOPT in Subsec. \ref{subsec:libraries}.
Given the \texttt{.uai} representing the BN and the \texttt{.uai.evid} random evidence:
\begin{verbatim}
 2
  4 1
  3 2
\end{verbatim}
DAOOPT would give the following output:
\begin{verbatim}
	--- Starting search ---
[0] u 3 4 -1.3581 5 2 1 2 2 1
[0] Cache statistics: . . . .

--------- Search done ---------
Problem name:  pomegranate
OR nodes:      3
AND nodes:     4
OR processed:  3
AND processed: 8
Leaf nodes:    2
Pruned nodes:  4
Deadend nodes: 1
Time elapsed:  0 seconds
Preprocessing: 0 seconds
-------------------------------
-1.3581 (0.0438433)

p 2 1 2
l 2 1 6
s -1.3581 5 2 1 2 2 1
\end{verbatim}
The end of the final line is the one of interest as it is the assignment of values to the variables that solves the MPE problem.
The \texttt{5 2 1 2 2 1} string is to be interpreted as meaning:
\begin{itemize}
  \item there are 5 variables in the solution
  \item the variable indexed by 0 (in the ordering given in the preable of the \texttt{.uai}) is assigned its second value (the ordering is inferred by the CPTs defined in the \texttt{.uai}) in the MPE solution
  \item variable 1 is assigned its second value
  \item variable 2 is assigned its third value
  \item variable 3 is assigned its third value
  \item variable 4 is assigned its second value
\end{itemize}
Variables 3 and 4 are constrained to assume the value specified in the input \texttt{.uai.evid}; in this case 1 and 2, respectively.

All the functionality relating to solving the MPE with DAOOPT is encapsulated in the \texttt{daoopt\_solver} function that given the input \texttt{.uai} files, returns the MPE solution.

\subsubsection{Other Machine Learning methods}
In order to quantify the prediction capabilities of the Bayesian Network, that is at the heart of this thesis' methods, I implemented a series of tests to find the best performing algorithm on the data set.
Given that the performance of Machine Learning algorithms is heavily dependent on the input classes, I used a process of \textit{exhaustive variable elimination} in order to identify the most relevant features for the predictions.
In parallel, I scored each input subset using the following ML algorithms, in order to find the best performing one:
\begin{itemize}
  \item Linear Regression: this method assumes that the relationship between the dependent variable $y$ and the regressors $x$ is linear i.e. that $y$ can be written as a linear combination of $x$'s components: $y = \beta_0 + \beta_1 x_1 + \ldots \beta_n x_n$.
  \item Logistic Regression: is used in lieu of Linear Regression when the values of the variables are categorical; it assumes that the relationship between the regressors $x$ and the log-odds of $y$ are linear i.e. $\log _{b} \frac{p}{1-p}=\beta_{0}+\beta_{1} x_{1}+ \ldots + \beta_{n} x_{n}$ with $p$ the probability of the event of interest.
  \item Linear Discriminant Analysis: LDA is related to Principal Component Analysis (PCA) in that it attempts to find a linear equation modelling the data but LDA explicitly tries to express the difference between the data classes.
  \item Decision Tree: a Decision Tree is built using a recursive, greedy algorithm that continually splits the dataset into two.  The variable along which to bisect is the one that yields the lowest accuracy loss in the resulting split.
  \item Na{\"i}ve Bayes: a Na{\"i}ve Bayes classifier is a conditional probability model that given features $x_1 \ldots x_n$, attempts to assign a probability to each of the possible outcomes $O_{k}$ of interest by using Bayes' Theorem (Eq. \ref{eq:bayes-theorem} ): $\mathbb{P}\left(O_{k} | x \right)=\frac{\mathbb{P}\left(O_{k}\right) \mathbb{P}\left(x| O_{k}\right)}{\mathbb{P}(x)}$.  The method is called \enquote{na{\"i}ve} because of a strong (ofter unrealistic) assumption of all the features $x_1 \ldots x_n$ being independent.
  \item K-Nearest Neighbours: the algorithm is non-parametric with the output class depending on the predominant class among the $k$ nearest neighbours (according to some distance metric) of the input vector $x$.
  \item Support-Vector: A Support-Vector Machine (SVM) is an algorithm that attempts to find the set of \textit{best-separating hyperplane} between classes of objects, seen as points in a high-dimensional space.  Such a hyperplane is the one that has maximum distance from the closest representatives of each class.
  \item Random Forest: this is an ensemble method that aims to correct Decision Trees' tendency to overfit the data.  A multitude of Decision Trees is constructed and the final classification output is the class that appears most often in the intermediate step.
  \item AdaBoost: Short for \textit{Adaptive Boosting}; this meta-learning, ensemble algorithm combines a series of weak classifiers, that may only be slight better than a random guess, through a weighted sum into a \textit{strong classifier}.  It is a meta-learning algorithm because the weak classifiers are revised over a series of iterations in order to improve their performance on previously misclassified instances.
  \end{itemize}
